Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2020-01-21 pldi14/sec-rw.tex
Revisions
Discuss lambda-diff WIP
Add lhs2TeX output for archival

% Emacs, this is -*- latex -*-!
\chapter{Related work}
\label{sec:rw}

@@

In the remainder of this section, we analyze the relation to the
most closely related prior works. \Citet{Ramalingam93}, \citet{Gupta99MMV}
and \citet{Acar06} discuss further related work.

@@

\section{Static approaches}
\pg{If we discuss partial evaluation, we should compare to
  \citep{Sundaresh91}}
Static approaches analyze a program at compile-time and produce an
incremental version that efficiently updates the output
of the original program according to changing inputs.

@@

\subsection{Finite differencing}
\label{sec:finite-diff}
\citet{Paige82FDC} present derivatives for a first-order language
with a fixed set of primitives.
\citet{Blakeley:1986:EUM} apply these ideas to a class of relational queries.
The database community extended
this work to queries on relational data, such as in \emph{algebraic
  differencing}~\citep{Gupta99MMV}, which inspired our work and
terminology. However, most of this work does not apply to nested
collections or algebraic data types, but only to relational
(flat) data, and no previous approach handles first-class
functions. Incremental support is typically designed
monolithically for a whole language, rather than piecewise.
Improving on algebraic differencing, \citet{Koch10IQE}
\emph{guarantees} asymptotic speedups with a compositional query
transformation and delivers huge speedup in realistic benchmarks,
though still for a first-order database language.

@@

>>>>

% Emacs, this is -*- latex -*-!
%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%
%
%
% First, let's redefine the forall, and the dot.
%
%
% This is made in such a way that after a forall, the next
% dot will be printed as a period, otherwise the formatting
% of `comp_` is used. By redefining `comp_`, as suitable
% composition operator can be chosen. Similarly, period_
% is used for the period.
%
\ReadOnlyOnce{forall.fmt}%
\makeatletter

% The HaskellResetHook is a list to which things can
% be added that reset the Haskell state to the beginning.
% This is to recover from states where the hacked intelligence
% is not sufficient.

\let\HaskellResetHook\empty
\newcommand*{\AtHaskellReset}[1]{%
  \g@addto@macro\HaskellResetHook{#1}}
\newcommand*{\HaskellReset}{\HaskellResetHook}

\global\let\hsforallread\empty

\newcommand\hsforall{\global\let\hsdot=\hsperiodonce}
\newcommand*\hsperiodonce[2]{#2\global\let\hsdot=\hscompose}
\newcommand*\hscompose[2]{#1}

\AtHaskellReset{\global\let\hsdot=\hscompose}

% In the beginning, we should reset Haskell once.
\HaskellReset

\makeatother
\EndFmtInput

% https://github.com/conal/talk-2015-essence-and-origins-of-frp/blob/master/mine.fmt
% Complexity notation:

% If an argument to a formatting directive starts with let, lhs2TeX likes to
% helpfully prepend a space to the let, even though that's seldom desirable.
% Write lett to prevent that.

% Hook into forall.fmt:
% Add proper spacing after forall-generated dots.

% We shouldn't use /=, that means not equal (even if it can be overriden)!

% XXX

%  format `stoup` = "\blackdiamond"

% Cancel the effect of \; (that is \thickspace)

% Use as in |vapply vf va (downto n) v|.
% (downto n) is parsed as an application argument, so we must undo the produced
% spacing.

% indexed big-step eval
% without environments
% big-step eval
% change big-step eval

% \, is 3mu, \! is -3mu, so this is almost \!\!.

\def\deriveDefCore{%
\begin{align*}
  \ensuremath{\Derive{\lambda (\Varid{x}\typcolon\sigma)\to \Varid{t}}} &= \ensuremath{\lambda (\Varid{x}\typcolon\sigma)\;(\Varid{dx}\typcolon\Delta \sigma)\to \Derive{\Varid{t}}} \\
  \ensuremath{\Derive{\Varid{s}\;\Varid{t}}} &= \ensuremath{\Derive{\Varid{s}}\;\Varid{t}\;\Derive{\Varid{t}}} \\
  \ensuremath{\Derive{\Varid{x}}} &= \ensuremath{\Varid{dx}} \\
  \ensuremath{\Derive{\Varid{c}}} &= \ensuremath{\DeriveConst{\Varid{c}}}
\end{align*}
}

% Drop unsightly numbers from function names. The ones at the end could be
% formatted as subscripts, but not the ones in the middle.

\chapter{Related work}
\label{sec:rw}
\label{ch:incr-rw}

@@

In the remainder of this section, we analyze the relation to the
most closely related prior works. \Citet{Ramalingam93}, \citet{Gupta99MMV}
and \citet{Acar06} discuss further related work.
Other related work, more closely to cache-transfer style, is
discussed in \cref{sec:cts-rw}.

@@

\section{Static approaches}
Static approaches analyze a program at compile-time and produce an
incremental version that efficiently updates the output
of the original program according to changing inputs.

@@

\citet{Sundaresh91} propose to incrementalize programs using
partial evaluation: given a partitioning of program inputs in parts
that change and parts that stay constant,
\citeauthor{Sundaresh91} partially evaluates a given program relative
to the constant input parts, and combine the result with the
changing inputs.

\subsection{Finite differencing}
\label{sec:finite-diff}
\citet{Paige82FDC} present derivatives for a first-order language
with a fixed set of primitives.
\citet{Blakeley:1986:EUM} apply these ideas to a class of relational queries.
The database community extended
this work to queries on relational data, such as in \emph{algebraic
  differencing}~\citep{Gupta99MMV}, which inspired our work and
terminology. However, most of this work does not apply to nested
collections or algebraic data types, but only to relational
(flat) data, and no previous approach handles first-class
functions or programs resulting from defunctionalization or
closure conversion. Incremental support is typically designed
monolithically for a whole language, rather than piecewise.
Improving on algebraic differencing, DBToaster (\citet{Koch10IQE,Koch14})
\emph{guarantees} asymptotic speedups with a compositional query
transformation and delivers huge speedup in realistic benchmarks,
though still for a first-order database language.
\pg{continue with new work, discuss why we don't do iterated differentiation.}

@@

It seems possible to transform higher-order functional programs
to database queries, using a variety of approaches
\citep{Grust:2009:FDP:1559845.1559982,Cheney2013practical}, some
of which support first-class functions via closure conversion
\citep{Grust2013first,Grust2013functions}, and incrementalize the
resulting programs using standard database technology. Such a
solution would inherit limitations of database incrementalization
approaches: in particular, it appears that database
incrementalization approaches such as DBToaster can handle the
insertion and removal of entire table rows, not of smaller
changes. Nevertheless, such an alternative approach might be
worth investigating.

Unlike later approaches to higher-order differentiation, we do
not restrict our base types to
groups unlike \citet{Koch2016incremental}, and transformed programs we
produce do not require further runtime
transformation unlike \citet{Koch2016incremental} and \citet{Huesca2015incrementality},
as we discuss further next.

\newcommand{\ldiff}{\TitleLambda--diff}
\subsection{\TitleLambda{}--diff and partial differentials}
\label{sec:rw-partial-differentials}
In work concurrent with \citet{CaiEtAl2014ILC}, \citet{Huesca2015incrementality}
introduces an alternative formalism, called \ldiff, for incremental computation
by program transformation. While \ldiff{} has some appealing features, it
currently appears to require program transformations at
runtime. Other systems appear to share this
feature~\citep{Koch2016incremental}. Hence, this section
discusses the reason in some detail.

Instead of differentiating a term \ensuremath{\Varid{t}} relative to all inputs (free
variables and function arguments) via \ensuremath{\Derive{\Varid{t}}}, like ILC,
\ldiff{} differentiates terms
relative to one input variable, and writes
$\sfrac{\partial t}{\partial x, d_x}$
for the result of differentiating $t$
relative to $x$, a term that computes the change in $t$ when the
value for $x$ is updated by change $d_x$. The formalism also uses
pointwise function changes, similarly to what we discussed in
\cref{ssec:pointwise-changes}.

Unfortunately, it is not known how to define such a
transformation for a $\lambda$-calculus with a standard
semantics, and the needed semantics appear to require runtime
term transformations, which are usually considered problematic
when implementing functional languages.
In particular, it appears necessary to introduce a new term
constructor \ensuremath{\Conid{D}\;\Varid{t}}, which evaluates \ensuremath{\Varid{t}} to a function value \ensuremath{\lambda \Varid{y}\to \Varid{u}}, and then evaluates to $\lambda (y, d_y) \to \sfrac{\partial
t}{\partial y, d_y}$, which differentiates \ensuremath{\Varid{t}} at runtime
relative to its head variable \ensuremath{\Varid{y}}. As an indirect consequence, if the
program under incrementalization contains function term \ensuremath{\Gamma\vdash\Varid{t}\typcolon\sigma\to \tau}, where \ensuremath{\Gamma} contains \ensuremath{\Varid{n}} free variables, it
can be necessary in the worst-case to differentiate \ensuremath{\Varid{t}} over any
subset of the \ensuremath{\Varid{n}} free variables in \ensuremath{\Gamma}. There are $2^n$ such
subsets. To perform all term transformations before runtime, it
seems hence necessary in the worst case to precompute $2^n$
partial derivatives for each function term \ensuremath{\Varid{t}}, which appears
unfeasible.
On the other hand, it is not clear how often this worst-case is
realized, or how big \ensuremath{\Varid{n}} grows in typical programs, or if it is
simply feasible to perform differentiation at runtime, similarly
to JIT compilation. Overall, an efficient implementation of
\ldiff{} remains an open problem.
It appears also \citet{Koch2016incremental}
suffer similar problems, but a few details appear simpler since
they restrict focus to functions over groups.

To see why \ldiff{} need introduce \ensuremath{\Conid{D}\;\Varid{t}}, consider
differentiating $\sfrac{\partial s\;t}{\partial x, d_x}$, that is,
the change $d$ of $s\;t$ when $x$x is updated by change $d_x$.
Change $d$ depends (a) on the change of $t$ when $x$ is updated
by $d_x$, that is
$\sfrac{\partial t}{\partial x, d_x}$;
(b) on how $s$ changes when its input $t$ is updated by
$\sfrac{\partial t}{\partial x, d_x}$; to express this change, \ldiff{}
expresses this via $\ensuremath{(\Conid{D}\;\Varid{s})\;\Varid{t}}\; \sfrac{\partial t}{\partial x, d_x}$;
(c) on the change of $s$ (applied to the updated $t$) when $x$ is
updated by $d_x$, that is $\sfrac{\partial t}{\partial x, d_x}$.
To compute component (b), \ldiff{} writes \ensuremath{\Conid{D}\;\Varid{s}} to
differentiate \ensuremath{\Varid{s}} not relative to \ensuremath{\Varid{x}}, but relative to the still unknown head
variable of \ensuremath{\Varid{s}}.
If \ensuremath{\Varid{s}} evaluates to \ensuremath{\lambda \Varid{y}\to \Varid{u}}, then \ensuremath{\Varid{y}} is the head variable of \ensuremath{\Varid{s}}, and \ensuremath{\Conid{D}\;\Varid{s}}
differentiates \ensuremath{\Varid{u}} relative to \ensuremath{\Varid{y}} and evaluates to $\lambda (y, d_y) \to
\sfrac{\partial u}{\partial y, d_y}$.
% Hence, \citeauthor{Huesca2015incrementality} adds term |D t| to
% the syntax of $\lambda$-calculus; evaluating

Overall, the rule for differentiating application in $\lambda$-diff is
\[
  \frac{\partial s\;t}{\partial x, d_x} = (D s)\left(t, \frac{\partial t}{\partial x,
d_x}\right) \circledcirc \frac{\partial s}{\partial x, d_x}\left(t \oplus \frac{\partial
t}{\partial x, d_x}\right).
  \]

This rule appears closely related to \cref{eq:pointwise-rewrite},
hence we refer to its discussion for clarification.

On the other hand, differentiating a term relative to all its
inputs introduces a different sort of overhead. For instance, it
is much more efficient to differentiate \ensuremath{\Varid{map}\;\Varid{f}\;\Varid{xs}} relative to
\ensuremath{\Varid{xs}} than relative to \ensuremath{\Varid{f}}: if \ensuremath{\Varid{f}} undergoes a non-nil change
\ensuremath{\Varid{df}}, \ensuremath{\Derive{\Varid{map}\;\Varid{f}\;\Varid{xs}}} must apply \ensuremath{\Varid{df}} to each elements in the
updated input \ensuremath{\Varid{xs}}. Therefore, in our practical implementations
\ensuremath{\Derive{\Varid{map}\;\Varid{f}\;\Varid{xs}}} tests whether \ensuremath{\Varid{df}} is nil and uses a more
efficient implementation. In \cref{sec:applying}, we detect at
compile time whether \ensuremath{\Varid{df}} is guaranteed to be nil. In
\cref{sec:incr-nest-loop}, we instead detect at runtime whether \ensuremath{\Varid{df}} is
nil. In both cases, authors of derivatives must implement this
optimization by hand. Instead, \ldiff{} hints at a more general
solution.
